The ability of single genes vs full genomes to resolve time and space in outbreak analysis

### 2019-10-15 genomic-horizon.tex
Reformat manuscript to BMC template, add final edits.

\documentclass[11pt,oneside,letterpaper]{article}
% \documentclass[9pt,lineno]{elife}

% graphicx package, useful for including eps and pdf graphics
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{subfig}
%\DeclareGraphicsExtensions{.pdf,.png,.jpg}

% basic packages
\usepackage{color}
\usepackage{parskip}
\usepackage{float}
\usepackage{microtype}
\usepackage{url}
\usepackage[T1]{fontenc}
\urlstyle{same}

% \setkeys{Gin}{draft} %% draft mode

% \usepackage[hidelinks]{hyperref}
% \hypersetup{colorlinks=true,linkcolor=black,citecolor=black,urlcolor=black}

\usepackage{longtable}
\usepackage[]{algorithm2e}

% reference figures across documents
\usepackage{xr}
%\externaldocument{mers-structure_supp}

% text layout
\usepackage{geometry}
\geometry{textwidth=15cm} % 15.25cm for single-space, 16.25cm for double-space
\geometry{textheight=22cm} % 22cm for single-space, 22.5cm for double-space

% helps to keep figures from being orphaned on a page by themselves
\renewcommand{\topfraction}{0.85}
\renewcommand{\textfraction}{0.1}

% bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,font=small]{caption}

% review layout with double-spacing
%\usepackage{setspace}
%\doublespacing
%\captionsetup{labelfont=bf,labelsep=period,font=doublespacing}

% cite package, to clean up citations in the main text. Do not remove.
%\usepackage{cite}
\usepackage{natbib}
%\renewcommand\citepleft{(}
%\renewcommand\citepright{)}
%\renewcommand\citepform[1]{\textsl{#1}}

\usepackage{amsmath}

%\usepackage{lineno}
%\linenumbers

% Remove brackets from numbering in list of References
%\renewcommand\refname{\large References}
%\makeatletter
%\renewcommand{\@biblabel}[1]{\quad#1.}
%\makeatother

\usepackage{authblk}
\renewcommand\Authands{ \& }
\renewcommand\Authfont{\normalsize \bf}
\renewcommand\Affilfont{\small \normalfont}
\makeatletter
\renewcommand\AB@affilsepx{, \protect\Affilfont}
\makeatother

% comments
%\usepackage{ulem}
\definecolor{purple}{rgb}{0.459,0.109,0.538}
\def\tb#1#2{\sout{#1} \textcolor{purple}{#2}}
\def\tbc#1{\textcolor{purple}{[#1]}}
\def\gdc#1{\textcolor{blue}{[#1]}}

% symbols
% \newcommand{\chiSq}{\chi^{2}_{df}} %LM: ancient DNA? =P %GD: quite :)
% \newcommand{\dtmrca}{\Delta_\mathrm{TMRCA}}
% \newcommand{\undtmrca}{\delta_\mathrm{TMRCA}}
% \newcommand{\dspr}{d_\mathrm{SPR}}

%%% Comments from lab meeting:
% - HIV in contour plot
% - time variation across a single mutation in RTT
% - review distances*probability
% - location inference - is the true location in the 95% set?
% - bias in collection impact on correlates of migration
% - mention that GLM is broad scale patterns and easy to get at
% - more focus on benefits of genomic epi
% - talk about impact of unsampled diversity i.e. whole clades missing
% - ancestry plots for all tips/summary across all tips. masked tip ancestry?
% - bar plot of mean waiting times between whole genomes and subgenomic fragments

%%% TITLE %%%
\title{\vspace{1.0cm} \LARGE \bf The ability of single genes vs full genomes to resolve time and space in outbreak analysis}

\author[1,2$\ast$]{Gytis Dudas}
\author[1]{Trevor Bedford}

\affil[1]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA, USA}
\affil[2]{Gothenburg Global Biodiversity Center, Gothenburg, Sweden}
\affil[$\ast$]{Corresponding author (gytisdudas@gmail.com)}
\begin{document}

\maketitle

\begin{abstract}
% Inexpensive pathogen genome sequencing has transformed the field of genetic epidemiology into genomic epidemiology.
Inexpensive pathogen genome sequencing has had a transformative effect on the field of phylodynamics, where ever increasing volumes of data have promised real-time insight into outbreaks of infectious disease.
As well as the sheer volume of pathogen isolates being sequenced, the sequencing of whole pathogen genomes, rather than select loci, has allowed phylogenetic analyses to be carried out at finer time scales, often approaching serial intervals for infections caused by rapidly evolving RNA viruses.
Despite its utility, whole genome sequencing of pathogens has not been adopted universally and targeted sequencing of loci is common in some pathogen-specific fields.
In this study we aim to highlight the utility of sequencing whole genomes of pathogens by re-analysing a well-characterised collection of Ebola virus sequences in the form of complete viral genomes ($\sim$19kb long) or the rapidly evolving glycoprotein (GP, $\sim$2kb long) gene.
We quantify changes in phylogenetic, temporal, and spatial inference resolution as a result of this reduction in data and compare these to theoretical expectations.
We propose a simple intuitive metric for quantifying temporal resolution, \textit{i.e.} the time scale over which sequence data might be informative of various processes as a quick back-of-the-envelope calculation of statistical power available to molecular clock analyses.
% We quantify the changes in phylogenetic, temporal, and spatial inference resolution available to researchers when complete virus genomes are used ($\sim$19kb long) versus just the rapidly evolving glycoprotein (GP) gene ($\sim$2kb long) of Ebola virus in West Africa.

% By quantifying errors in phylogenetic reconstruction we hope to persuade researchers looking to sequence pathogens during outbreaks of the utility of complete genome sequencing.
% We further generalise our results by relating four parameters - probability, evolutionary rate, alignment length and waiting times - that determine how much information can be encoded by evolving pathogen lineages via mutations, which can help guide future sequencing studies.
% \tbc{The use of `probability', etc... here is a bit opaque.'}
\end{abstract}

\pagebreak

\section*{Introduction}
The combination of decreasing cost of sequencing and the unparalleled insight it offers have led to the adoption of pathogen genetic sequencing as one of the most effective tools in a modern epidemiologist's toolkit.
When coupled with sophisticated models of evolution, pathogen sequences can be used to look into epidemiological features such as cryptic transmission \citep{faria_establishment_2017}, migration \citep{lemey_bayesian_2009,lemey_unifying_2014}, and origins \citep{smith_origins_2009} of infectious diseases, amongst others.
Pathogen sequences also contain information about past temporal dynamics before sequence data have been collected \citep{raghwani_origin_2012} due to the pattern of shared and unique mutations inherited from preceding generations.
Molecular phylogenetic approaches rely on decoding these patterns of shared mutations into a nested graph known as the phylogenetic tree.
Pathogens often have short generation times and some, like RNA viruses, also possess polymerases with low replication fidelity such that mutations are generated at a rapid pace \citep{drummond_measurably_2003,biek_measurably_2015} leading to fast differentiation of pathogen lineages at the genetic level as they spread.
With appropriate sampling and information (``metadata'') about sequences, historic population dynamics can be inferred and quantified from pathogen phylogenies.
Changes in pathogen population sizes over time \citep{pybus_integrated_2000}, inference of unobserved ancestral states \citep{lemey_bayesian_2009,dudas_mers-cov_2018}, correlates of processes \citep{faria_simultaneously_2013,lemey_unifying_2014,dudas_virus_2017}, and overall phylodynamic \citep{grenfell_unifying_2004} patterns can be inferred from molecular phylogenies and used to understand patterns of pathogen transmission at a number of scales.

Before widespread adoption of high-throughput sequencing limitations and costs led to amplification and sequencing of short fragments of pathogen genomes \citep{jin_genetic_1999,jin_proposal_2005}.
These subgenomic fragments were often chosen for their diversity, such as viral surface glycoproteins that experience selective pressures from vertebrate immune systems, or their utility, such as routine sequencing of HIV \textit{pol} gene to test for drug resistance \citep{kaye_phylogenetic_2008,rhee_human_2003}.
Whilst subgenomic fragments of pathogens are very accurate and specific as diagnostic markers and informative about long-term evolution, their length (dictated by the compromise between information content and ease of sequencing) limits their utility in detailed molecular epidemiology investigations, for example during outbreaks \citep{wohl_co-circulating_2018}, as only mutations occurring within the small region of the genome that is sequenced are available for phylogenetic inference.

Molecular clocks have been particularly useful in molecular epidemiology, where the accumulation of mutations between sequences is used as a noisy approximation for elapsed time, given either times of events in the phylogeny (sequence dates or dates of common ancestors) or a previously determined molecular clock rate.
Generally, neutral pathogen variation at the nucleotide level ebbs and flows under the forces of population genetics, unlike beneficial or deleterious variation which tends to either fix or be purged rapidly, respectively.
Due to their random and discrete nature, mutations are modelled as a Poisson process \citep{yang_computational_2006} where the waiting time $t$ for observing a mutation at a single site is exponentially distributed with evolutionary rate parameter $R$.
The probability of observing 0 mutations at a single site after time $t$ is $e^{-Rt}$ and the probability of at least one mutation is therefore $1-e^{-Rt}$.
Higher evolutionary rates \textit{R} or waiting times \textit{t} result in higher probabilities of observing at least one mutation at the site in question.
Since sites are assumed to evolve independently, the probability of observing at least one mutation across \textit{L} sites is
\begin{equation}
  P = 1-e^{-RLt},
\end{equation}
where \textit{RL} is expressed in substitutions per year (rate in substitutions per site per year multiplied by number of sites).
Since the probability of observing at least one mutation changes depending on waiting time (and \textit{vice versa}) we instead focus on the mean waiting time until at least one mutation appears.
The mean of an exponential distribution is $\lambda^{-1}$ where $\lambda$ in our case is \textit{RL}, such that mean waiting time $\bar{t}$ under a given evolutionary rate \textit{R}, and sequence length \textit{L} becomes
\begin{equation}
  \bar{t} = \frac{1}{RL}
  \label{horizon}
\end{equation}
When the evolutionary rate \textit{R} or sequence length \textit{L} are low mean waiting times $\bar{t}$ are lengthened and \textit{vice versa}.
It also suggests a worrying relationship between \textit{R}, \textit{L}, and $\bar{t}$ - a reduction in either \textit{R} or \textit{L} leads to a reciprocal reduction in $\bar{t}$, such that reducing the number of sites by 10\%, for example, requires a $\frac{1}{0.9} = 1.11(1)$-fold increase in evolutionary rate to maintain the same temporal resolution $\bar{t}$ or risk increasing the waiting time for a mutation by the same amount (\textit{i.e.} $\approx$11\%, or 0.9 of the original resolution available).
This gets worse reciprocally, such that halving the sites requires doubling the evolutionary rate, using 10\% of the sites requires a ten-fold increase in evolutionary rate, \textit{etc}.

Since both maximum plausible evolutionary rates $R$ and genome length $G$ are largely dictated by deleterious mutation load \citep{gago_extremely_2009}, neither quantity will vary substantially for a given pathogen, though individually $R$ and $G$ can vary substantially, where for example viruses have high $R$ and low $G$ on average, and bacteria have higher $G$ but lower $R$.
Sequencing recovers some fraction $f$ of the genome length $G$ ($L=Gf$) for analyses, and sequencing complete genomes ($f=1.0; L = G$) is the best possible scenario, since sequencing any shorter region requires the evolutionary rate to increase by a factor of $\frac{1}{f}$, which even if $L = 0.4 G$ means the evolutionary rate would have to be 2.5 times faster in the remaining 0.4 of the genome to be able to record information in the form of mutations at the same speed as complete genomes.
The message of our manuscript, at least as far as densely sampled infectious disease outbreaks go, is that the task of sequencing a complete pathogen genome will rarely be as miserable a task as analysing a fraction of one.

In this study we show this by quantifying how much information relevant to phylodynamic analysis is lost when shorter genomic regions are used instead of full genomes.
By focusing our attention on a subset (600 sequences) of a well-characterised genomic sequence data (comprised of >1600 viral genomes) set derived from the West African Ebola virus epidemic of 2013-2015 \citep{dudas_virus_2017} we estimate loss in precision and accuracy of molecular clock models and phylogeographic inference methods when only the glycoprotein gene (GP), a region representing just 10\% of the viral genome, is analysed despite GP evolving at rates faster than the genomic average.
Our methods rely on masking tip dates and locations for 60 (10\%) of the sequences in a classic training-testing split, where we re-infer these parameters as latent variables using Markov chain Monte Carlo (MCMC).
We show that this reduction in data not only leads to severe mixing issues in MCMC analyses by removing the constraints additional data impose on plausible parameter space without adding restrictive priors to compensate, but can also result in unreliable tip date and location inference.
Despite achieving much better temporal resolution when using complete viral genomes we still find residual error caused by inherent randomness of mutations which is close to theoretical expectations (Eq \ref{horizon}).
We refer to this as the temporal horizon, \textit{i.e.} a temporal resolution limit where population processes occurring at a rate faster than the rate at which mutations enter and are observed in a population will not be captured with high fidelity, even with genome sequences.

\begin{figure}[h]
 \centering
	\includegraphics[width=0.75\textwidth]{figures/fig1_trees.png}
	\caption{\textbf{Phylogenies of West African Ebola virus genomes (top) or GP sequences (bottom).}
	Temporal phylogenies recovered using BEAST are shown on the left, maximum likelihood phylogenies recovered with RAxML are on the right.
  Tips are coloured based on country (Sierra Leone in blue, Liberia in red, Guinea in green) and location (lighter colours indicate administrative divisions lying towards west of the country).
  Tips outlined in white indicate the 60 chosen for date and location masking, ticks to the right of phylogenies indicate the y positions of masked tips, coloured by their true location.
  In temporal phylogenies branches are also coloured based on GLM-inferred ancestral locations.
  Nodes in temporal phylogenies with <0.10 posterior probability are indicated with grey X marks.
  Maximum likelihood phylogenies on the right are rooted via temporal regression in TreeTime.
	}
	\label{trees}
\end{figure}

@@

\subsection*{Loss of phylogenetic signal}
% In the past amplifying smaller fragments of pathogen genomes was often a pragmatic compromise between the cost of sequencing and information content, leading to sequencing of short but rapidly evolving pathogen genes or genomic regions.
% What few sequences were available were often collected months apart or across large geographic areas, ensuring that many sequences were genetically distinct either via mutations occurring in the intervening time or through prior population differentiation.
% However, in outbreak scenarios shorter genome fragments often do not accumulate mutations at a pace rapid enough to be of much use over short timescales.

Figure \ref{trees} shows the reconstructed phylogenies in substitution space (right) and time space (left) for 600 complete Ebola virus genomes (top) or just GP sequences (bottom).
Although higher levels of divergence are observed in the GP dataset (note that x axes between genome and GP phylogenies are not consistent), as seen from tree height, the differences in the number of non-polytomic nodes between genomic and GP data are clear, indicating substantially better resolution in disentangling the exact relationships between lineages in the former.
Figure \ref{embedding} shows where in the better resolved maximum likelihood phylogeny of genome sequences mutations that occurred in just the region spanning the GP gene.
Internal branches of a phylogeny correspond to hypotheses of common ancestry and in the case of GP only 42 internal nodes are identified in the maximum likelihood phylogeny compared to 210 internal nodes for complete genomes.
The few aspects of the West African epidemic that can be inferred from both GP and genome phylogenies, only the virus' origins in Guinea are clear, with details of its onwards spread largely lost in the GP phylogeny.
Genomic data, on the other hand, despite a reduction from over 1600 sequences described in the original study \citep{dudas_virus_2017} down to just 600, still contain information about the role of Sierra Leone's epidemic in maintaining transmission across the region through both endemic proliferation of lineages and their spread to neighbouring countries.

@@

In contrast to the maximum likelihood phylogeny of GP on the right (Figure \ref{trees}) its corresponding temporal phylogeny on the left exhibits a reconstruction of the West African epidemic largely consistent with what has been established previously \citep{dudas_virus_2017}.
This is likely to be caused by the combined effects of two sources of information.
First, additional information is added by specifying the collection dates for sequences, which might exclude certain topologies from being considered during MCMC on account of the relatively small effective population size of Ebola virus in West Africa.
Second, the generalised linear model approach to inferring migration is information-rich as it provides over 3000 possible parameter values (pairwise migration rates between locations) per predictor matrix, and thus if a few branches are strongly selecting for a ``correct'' predictor matrix to be included in the migration model that predictor matrix can then be used to determine the likely locations of branches for which less information is available.
However, a simpler continuous time Markov chain model where each individual pairwise migration rate is inferred individually in a maximum likelihood framework exhibits broadly similar patterns too (Figure \ref{TTtrees}).

>>>>

%% BioMed_Central_Tex_Template_v1.06
%%                                      %
%  bmc_article.tex            ver: 1.06 %
%                                       %

%%IMPORTANT: do not delete the first line of this template
%%It must be present to enable the BMC Submission system to
%%recognise this template!!

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                     %%
%%  LaTeX template for BioMed Central  %%
%%     journal article submissions     %%
%%                                     %%
%%          <8 June 2012>              %%
%%                                     %%
%%                                     %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% For instructions on how to fill out this Tex template           %%
%% document please refer to Readme.html and the instructions for   %%
%% authors page on the biomed central website                      %%
%% http://www.biomedcentral.com/info/authors/                      %%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%% BioMed Central currently use the MikTex distribution of         %%
%% TeX for Windows) of TeX and LaTeX.  This is available from      %%
%% http://www.miktex.org                                           %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% additional documentclass options:
%  [doublespacing]
%  [linenumbers]   - put the line numbers on margins

%%% loading packages, author definitions

%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
\documentclass{bmcart}

%%% Load packages
%\usepackage{amsthm,amsmath}
%\RequirePackage{natbib}
%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
%\RequirePackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
% \usepackage{lmodern}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                             %%
%%  If you wish to display your graphics for   %%
%%  your own use using includegraphic or       %%
%%  includegraphics, then comment out the      %%
%%  following two lines of code.               %%
%%  NB: These line *must* be included when     %%
%%  submitting to BMC.                         %%
%%  All figure files must be submitted as      %%
%%  separate graphics through the BMC          %%
%%  submission process, not included in the    %%
%%  submitted article.                         %%
%%                                             %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\def\includegraphic{}
\def\includegraphics{}

%%% Put your definitions there:
\startlocaldefs
\endlocaldefs

%%% Begin ...
\begin{document}

%%% Start of article front matter
\begin{frontmatter}

\begin{fmbox}
\dochead{Research}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the title of your article here     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{The ability of single genes vs full genomes to resolve time and space in outbreak analysis}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors here                   %%
%%                                          %%
%% Specify information, if available,       %%
%% in the form:                             %%
%%   <key>={<id1>,<id2>}                    %%
%%   <key>=                                 %%
%% Comment or delete the keys which are     %%
%% not used. Repeat \author command as much %%
%% as required.                             %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\author[
   addressref={aff1,aff2},                   % id's of addresses, e.g. {aff1,aff2}
   corref={aff1},                       % id of corresponding address, if any
   % noteref={n1},                        % id's of article notes, if any
   email={gytisdudas@gmail.com}   % email address
]{\inits{GD}\fnm{Gytis} \snm{Dudas}}
\author[
   addressref={aff1},
   email={trevor@bedford.io}
]{\inits{TB}\fnm{Trevor} \snm{Bedford}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors' addresses here        %%
%%                                          %%
%% Repeat \address commands as much as      %%
%% required.                                %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\address[id=aff1]{%                           % unique id
  \orgname{Fred Hutchinson Cancer Research Center}, % university, etc
  \street{1100 Fairview Ave N},                     %
  \postcode{98109}                                % post or zip code
  \city{Seattle},                              % city
  \cny{USA}                                    % country
}
\address[id=aff2]{%
  \orgname{Gothenburg Global Biodiversity Centre},
  \street{Carl Skottsbergs gata 22B},
  \postcode{413 19}
  \city{Gothenburg},
  \cny{Sweden}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter short notes here                   %%
%%                                          %%
%% Short notes will be after addresses      %%
%% on first page.                           %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{artnotes}
%\note{Sample of title note}     % note to the article
% \note[id=n1]{Equal contributor} % note, connected to author
\end{artnotes}

\end{fmbox}% comment this for two column layout

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Abstract begins here                 %%
%%                                          %%
%% Please refer to the Instructions for     %%
%% authors on http://www.biomedcentral.com  %%
%% and include the section headings         %%
%% accordingly for your article type.       %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{abstractbox}

\begin{abstract} % abstract

\parttitle{Background}
Inexpensive pathogen genome sequencing has had a transformative effect on the field of phylodynamics, where ever increasing volumes of data have promised real-time insight into outbreaks of infectious disease.
As well as the sheer volume of pathogen isolates being sequenced, the sequencing of whole pathogen genomes, rather than select loci, has allowed phylogenetic analyses to be carried out at finer time scales, often approaching serial intervals for infections caused by rapidly evolving RNA viruses.
Despite its utility, whole genome sequencing of pathogens has not been adopted universally and targeted sequencing of loci is common in some pathogen-specific fields.

\parttitle{Results}
In this study we highlighted the utility of sequencing whole genomes of pathogens by re-analysing a well-characterised collection of Ebola virus sequences in the form of complete viral genomes ($\approx$19kb long) or the rapidly evolving glycoprotein (GP, $\approx$2kb long) gene.
We have quantified changes in phylogenetic, temporal, and spatial inference resolution as a result of this reduction in data and compared these to theoretical expectations.

\parttitle{Conclusions}
We propose a simple intuitive metric for quantifying temporal resolution, \textit{i.e.} the time scale over which sequence data might be informative of various processes as a quick back-of-the-envelope calculation of statistical power available to molecular clock analyses.

\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The keywords begin here                  %%
%%                                          %%
%% Put each keyword in separate \kwd{}.     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{keyword}
\kwd{phylogenetics}
\kwd{phylogeography}
\kwd{genomic epidemiology}
\kwd{Ebola virus}
\end{keyword}

% MSC classifications codes, if any
%\begin{keyword}[class=AMS]
%\kwd[Primary ]{}
%\kwd{}
%\kwd[; secondary ]{}
%\end{keyword}

\end{abstractbox}
%
%\end{fmbox}% uncomment this for twcolumn layout

\end{frontmatter}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% The Main Body begins here                %%
%%                                          %%
%% Please refer to the instructions for     %%
%% authors on:                              %%
%% http://www.biomedcentral.com/info/authors%%
%% and include the section headings         %%
%% accordingly for your article type.       %%
%%                                          %%
%% See the Results and Discussion section   %%
%% for details on how to create sub-sections%%
%%                                          %%
%% use \cite{...} to cite references        %%
%%  \cite{koon} and                         %%
%%  \cite{oreg,khar,zvai,xjon,schn,pond}    %%
%%  \nocite{smith,marg,hunn,advi,koha,mouse}%%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%% start of article main body
% <put your article body there>

%%%%%%%%%%%%%%%%
%% Background %%
%%
%
%
\section*{Introduction}

The combination of decreasing cost of sequencing and the unparalleled insight it offers have led to the adoption of pathogen genetic sequencing as one of the most effective tools in a modern epidemiologist's toolkit.
When coupled with sophisticated models of evolution, pathogen sequences can be used to look into epidemiological features such as cryptic transmission \cite{faria_establishment_2017}, migration \cite{lemey_bayesian_2009,lemey_unifying_2014}, and origins \cite{smith_origins_2009} of infectious diseases, amongst others.
Pathogen sequences also contain information about past temporal dynamics before sequence data have been collected \cite{raghwani_origin_2012} due to the pattern of shared and unique mutations inherited from preceding generations.
Molecular phylogenetic approaches rely on decoding these patterns of shared mutations into a nested graph known as the phylogenetic tree.
Pathogens often have short generation times and some, like RNA viruses, also possess polymerases with low replication fidelity such that mutations are generated at a rapid pace \cite{drummond_measurably_2003,biek_measurably_2015} leading to fast differentiation of pathogen lineages at the genetic level as they spread.
With appropriate sampling and information (``metadata'') about sequences, historic population dynamics can be inferred and quantified from pathogen phylogenies.
Changes in pathogen population sizes over time \cite{pybus_integrated_2000}, inference of unobserved ancestral states \cite{lemey_bayesian_2009,dudas_mers-cov_2018}, correlates of processes \cite{faria_simultaneously_2013,lemey_unifying_2014,dudas_virus_2017}, and overall phylodynamic \cite{grenfell_unifying_2004} patterns can be inferred from molecular phylogenies and used to understand patterns of pathogen transmission at a number of scales.

Before widespread adoption of high-throughput sequencing limitations and costs led to amplification and sequencing of short fragments of pathogen genomes \cite{jin_genetic_1999,jin_proposal_2005}.
These subgenomic fragments were often chosen for their diversity, such as viral surface glycoproteins that experience selective pressures from vertebrate immune systems, or their utility, such as routine sequencing of HIV \textit{pol} gene to test for drug resistance \cite{kaye_phylogenetic_2008,rhee_human_2003}.
Whilst subgenomic fragments of pathogens are very accurate and specific as diagnostic markers and informative about long-term evolution, their length (dictated by the compromise between information content and ease of sequencing) limits their utility in detailed molecular epidemiology investigations, for example during outbreaks \cite{wohl_co-circulating_2018}, as only mutations occurring within the small region of the genome that is sequenced are available for phylogenetic inference.

Molecular clocks have been particularly useful in molecular epidemiology, where the accumulation of mutations between sequences is used as a noisy approximation for elapsed time, given either times of events in the phylogeny (sequence dates or dates of common ancestors) or a previously determined molecular clock rate.
Generally, neutral pathogen variation at the nucleotide level ebbs and flows under the forces of population genetics, unlike beneficial or deleterious variation which tends to either fix or be purged rapidly, respectively.
Due to their random and discrete nature, mutations are modelled as a Poisson process \cite{yang_computational_2006} where the waiting time $t$ for observing a mutation at a single site is exponentially distributed with evolutionary rate parameter $R$.
The probability of observing 0 mutations at a single site after time $t$ is $e^{-Rt}$ and the probability of at least one mutation is therefore $1-e^{-Rt}$.
Higher evolutionary rates \textit{R} or waiting times \textit{t} result in higher probabilities of observing at least one mutation at the site in question.
Since sites are assumed to evolve independently, the probability of observing at least one mutation across \textit{L} sites is
\begin{equation}
  P = 1-e^{-RLt},
\end{equation}
where \textit{RL} is expressed in substitutions per year (rate in substitutions per site per year multiplied by number of sites).
Since the probability of observing at least one mutation changes depending on waiting time (and \textit{vice versa}) we instead focus on the mean waiting time until at least one mutation appears.
The mean of an exponential distribution is $\lambda^{-1}$ where $\lambda$ in our case is \textit{RL}, such that mean waiting time $\bar{t}$ under a given evolutionary rate \textit{R}, and sequence length \textit{L} becomes
\begin{equation}
  \bar{t} = \frac{1}{RL}
  \label{horizon}
\end{equation}
When the evolutionary rate \textit{R} or sequence length \textit{L} are low mean waiting times $\bar{t}$ are lengthened and \textit{vice versa}.
It also suggests a worrying relationship between \textit{R}, \textit{L}, and $\bar{t}$ - a reduction in either \textit{R} or \textit{L} leads to a reciprocal reduction in $\bar{t}$, such that reducing the number of sites by 10\%, for example, requires a $\frac{1}{0.9} = 1.11(1)$-fold increase in evolutionary rate to maintain the same temporal resolution $\bar{t}$ or risk increasing the waiting time for a mutation by the same amount (\textit{i.e.} $\approx$11\%, or 0.9 of the original resolution available).
This gets worse reciprocally, such that halving the sites requires doubling the evolutionary rate, using 10\% of the sites requires a ten-fold increase in evolutionary rate, \textit{etc}.

Since both maximum plausible evolutionary rates $R$ and genome length $G$ are largely dictated by deleterious mutation load \cite{gago_extremely_2009}, neither quantity will vary substantially for a given pathogen, though individually $R$ and $G$ can vary substantially, where for example viruses have high $R$ and low $G$ on average, and bacteria have higher $G$ but lower $R$.
Sequencing recovers some fraction $f$ of the genome length $G$ ($L=Gf$) for analyses, and sequencing complete genomes ($f=1.0; L = G$) is the best possible scenario, since sequencing any shorter region requires the evolutionary rate to increase by a factor of $\frac{1}{f}$, which even if $L = 0.4 G$ means the evolutionary rate would have to be 2.5 times faster in the remaining 0.4 of the genome to be able to record information in the form of mutations at the same speed as complete genomes.
The message of our manuscript, at least as far as densely sampled infectious disease outbreaks go, is that the task of sequencing a complete pathogen genome will rarely be as miserable a task as analysing a fraction of one.

In this study we show this by quantifying how much information relevant to phylodynamic analysis is lost when shorter genomic regions are used instead of full genomes.
By focusing our attention on a subset (600 sequences) of a well-characterised genomic sequence data (comprised of $>$1600 viral genomes) set derived from the West African Ebola virus epidemic of 2013-2015 \cite{dudas_virus_2017} we estimate loss in precision and accuracy of molecular clock models and phylogeographic inference methods when only the glycoprotein gene (GP), a region representing just 10\% of the viral genome, is analysed despite GP evolving at rates faster than the genomic average.
Our methods rely on masking tip dates and locations for 60 (10\%) of the sequences in a classic training-testing split, where we re-infer these parameters as latent variables using Markov chain Monte Carlo (MCMC).
We show that this reduction in data not only leads to severe mixing issues in MCMC analyses by removing the constraints additional data impose on plausible parameter space without adding restrictive priors to compensate, but can also result in unreliable tip date and location inference.
Despite achieving much better temporal resolution when using complete viral genomes we still find residual error caused by inherent randomness of mutations which is close to theoretical expectations (Eq \ref{horizon}).
We refer to this as the temporal horizon, \textit{i.e.} a temporal resolution limit where population processes occurring at a rate faster than the rate at which mutations enter and are observed in a population will not be captured with high fidelity, even with genome sequences.

@@

\subsection*{Loss of phylogenetic signal}

Figure \ref{trees} shows the reconstructed phylogenies in substitution space (right) and time space (left) for 600 complete Ebola virus genomes (top) or just GP sequences (bottom).
Although higher levels of divergence are observed in the GP dataset, as seen from tree height, the differences in the number of non-polytomic nodes between genomic and GP data are clear, indicating substantially better resolution in disentangling the exact relationships between lineages in the former.
Figure \ref{embedding} shows where in the better resolved maximum likelihood phylogeny of genome sequences the mutations that occurred in just the region spanning the GP gene are located, with continuous blocks of colours corresponding to regions of the tree that would become collapsed if only the GP data were used and the tree were inferred correctly.
Internal branches of a phylogeny correspond to hypotheses of common ancestry and in the case of GP only 42 internal nodes are identified in the maximum likelihood phylogeny compared to 210 internal nodes for complete genomes.
The one aspect of the West African epidemic that can be inferred from both GP and genome phylogenies is that the virus' origins lie in Guinea, but details of its onwards spread are largely lost in the GP phylogeny.
Genomic data, on the other hand, despite a reduction from over 1600 sequences described in the original study \cite{dudas_virus_2017} down to just 600, still contain information about the role of Sierra Leone's epidemic in maintaining transmission across the region through both endemic proliferation of lineages and their spread to neighbouring countries.

@@

In contrast to the maximum likelihood phylogeny of GP on the right (Figure \ref{trees}) its corresponding temporal phylogeny on the left exhibits a reconstruction of the West African epidemic largely consistent with what has been established previously \cite{dudas_virus_2017}.
This is likely to be caused by the combined effects of two sources of information.
First, additional information is added by specifying the collection dates for sequences, which might exclude certain topologies from being considered during MCMC on account of the relatively small effective population size of Ebola virus in West Africa.
Second, the generalised linear model approach to inferring migration is information-rich as it provides over 3000 possible parameter values (pairwise migration rates between locations) per predictor matrix, and thus if a few branches are strongly selecting for a ``correct'' predictor matrix to be included in the migration model that predictor matrix can then be used to determine the likely locations of branches for which less information is available.
However, a simpler continuous time Markov chain model where each individual pairwise migration rate is inferred individually in a maximum likelihood framework exhibits broadly similar patterns too (Figure \ref{TTtrees}).
